Program Optimisations via Hylomorphisms for Extraction of Executable Code

### 2025-07-07 paper/paper.tex
Hanging words

For example, the category of sets, denoted by $\Set$, is the category where
objects are sets and arrows are functions between sets. The initial object $0$
in $\Set$ is the empty set $\emptyset$ and the terminal object $1$ is any
singleton set. The reader who is not accustomed with category theory can assume
types are sets and programs are functions, giving the intuition that the category $\Set$ can also be viewed as the category of (simple) types and programs between them.

@@

For example, the terminal coalgebra for the functor $BX = A \times X$ is the set
of infinite streams over the set $A$, that is the greatest solution to the equation
$\Str{A} \iso A \times \Str{A}$.  The uniqueness property of terminal
$B$-coalgebras states that for any other $B$-coalgebra $(X, c_{X})$ there exists
a unique $B$-coalgebra homomorphism into the terminal coalgebra
$(\nu B, \OutIso)$ which is denoted by $\ana{c_{X}}$ and pronounced
``\emph{anamorphism}'' or ``\emph{unfold}''. We spell out the uniqueness
property:
\begin{equation}
  \label{eq:ana:unique}
  f = \ana{c_{X}} \iff f = \OutOp \circ B(f) \circ c_{X}
\end{equation}
From the uniqueness property we can derive the fusion law for unfolds:
\begin{equation}
  \label{eq:ana:fusion}
  \ana{c_{Y}} \circ f = \ana{c_{X}}
\end{equation}
for all $B$-coalgebra homomorphisms $f : (X, c_{X}) \to (Y, c_{Y})$.

@@

The \coq{LFix F} setoid equality is the least fixed point of the \coq{App F}
setoid equality, i.e. if \coq{LFixR F} represents the setoid equality of
\coq{LFix F}, and we have two \coq{x y : LFix F}, then \coq{LFixR F x y} iff
\begin{coqcode}
shape x =e shape y /\ (forall p p', projT1 p = projT1 p'
   -> LFixR F (cont x p) (cont y p'))
\end{coqcode}
We define smart constructors for the isomorphism of least fixed
points as proper morphisms:
\begin{coqcode}
l_in : App F (LFix F) ~> LFix F         l_out : LFix F ~> App F (LFix F)
\end{coqcode}
Catamorphisms are constructed so that they structurally deconstruct the
datatype, call themselves recursively, and then compose the result using an
\coq{F}-algebra.
\begin{coqcode}
Definition cata_f (a : Alg F A) : LFix F -> A
  := fix f (x : LFix) := match x with
        | LFix_in x => a {shape := shape x; cont := fun e => f (cont x e)}
        end.
\end{coqcode}
It is easy to prove that catamorphisms respect setoid equalities.  We
define \coq{cata} as a setoid morphism between the setoid of \coq{F}-algebras and the
setoid of functions from \coq{LFix F} to \coq{A}:
\begin{coqcode}
cata : forall `{setoid A}, Alg F A ~> (LFix F ~> A)
\end{coqcode}
Finally, we prove that catamorphisms satisfy their universal property (see
Section~\ref{sec:category-theory}):
\begin{coqcode}
Lemma cata_univ `{eA : setoid A} (alg : Alg F A) (f : LFix ~> A)
    : f =e cata alg <-> f =e alg \o fmap f \o l_out.
\end{coqcode}
In other words, if there is any other \coq{f} with the same structural recursive
shape as the catamorphism on the algebra \coq{alg} then it must be equal to that
catamorphism.

>>>>

For example, the category of sets, denoted by $\Set$, is the category where
objects are sets and arrows are functions between sets. The initial object $0$
in $\Set$ is the empty set $\emptyset$ and the terminal object $1$ is any
singleton set. The reader who is not accustomed with category theory can assume
for simplicity that  types may be viewed as sets and programs as functions
between sets, giving the intuition that the category $\Set$ can also be viewed
as the category of (simple) types and programs between them.

@@

For example, the terminal coalgebra for the functor $BX = A \times X$ is the set
of infinite streams over the set $A$, that is the greatest solution to the equation
$\Str{A} \iso A \times \Str{A}$.  The uniqueness property of terminal
$B$-coalgebras states that for any other $B$-coalgebra $(X, c_{X})$ there exists
a unique $B$-coalgebra homomorphism into the terminal coalgebra
$(\nu B, \OutIso)$ which is denoted by $\ana{c_{X}}$ and pronounced
``\emph{anamorphism}'' or ``\emph{unfold}''. We spell out the uniqueness
property of unfolds:
\begin{equation}
  \label{eq:ana:unique}
  f = \ana{c_{X}} \iff f = \OutOp \circ B(f) \circ c_{X}
\end{equation}
From the uniqueness property we can derive the fusion law for unfolds:
\begin{equation}
  \label{eq:ana:fusion}
  \ana{c_{Y}} \circ f = \ana{c_{X}}
\end{equation}
for all $B$-coalgebra homomorphisms $f : (X, c_{X}) \to (Y, c_{Y})$.

@@

The \coq{LFix F} setoid equality is the least fixed point of the \coq{App F}
setoid equality, i.e. if \coq{LFixR F} represents the setoid equality of
\coq{LFix F}. Let \coq{x y : LFix F}, then \coq{LFixR F x y} iff
\begin{coqcode}
shape x =e shape y /\ (forall p p', projT1 p = projT1 p'
   -> LFixR F (cont x p) (cont y p'))
\end{coqcode}
We define smart constructors for the isomorphism of least fixed
points as proper morphisms:
\begin{coqcode}
l_in : App F (LFix F) ~> LFix F         l_out : LFix F ~> App F (LFix F)
\end{coqcode}
Catamorphisms are constructed so that they structurally deconstruct the
datatype, call themselves recursively, and then compose the result using an
\coq{F}-algebra.
\begin{coqcode}
Definition cata_f (a : Alg F A) : LFix F -> A
  := fix f (x : LFix) := match x with
        | LFix_in x => a {shape := shape x; cont := fun e => f (cont x e)}
        end.
\end{coqcode}
It is easy to prove that catamorphisms respect setoid equalities.  We
define \coq{cata} as a setoid morphism between the setoid of \coq{F}-algebras and the
setoid of functions from \coq{LFix F} to \coq{A}:
\begin{coqcode}
cata : forall `{setoid A}, Alg F A ~> (LFix F ~> A)
\end{coqcode}
Finally, we prove that catamorphisms satisfy their universal property (see
Section~\ref{sec:category-theory}):
\begin{coqcode}
Lemma cata_univ `{eA : setoid A} (alg : Alg F A) (f : LFix ~> A)
    : f =e cata alg <-> f =e alg \o fmap f \o l_out.
\end{coqcode}
In other words, if there is any other \coq{f} with the same structural recursive
shape as the catamorphism on the algebra \coq{alg} then it must be equal to that
catamorphism.
